---
title: "Stat5361-Homework 3"
author: "Xueying Li & Huiqiong Liao"
date: "2/25/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

##Question1 
####Derive the updating rules in the given algorithm based on the construction of an EM algorithm.   

$$Q[\Psi|\Psi(k)]=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}[log(\pi_i)+log(\phi(y_i-x_i^T \beta_j;0,\sigma^2))]$$   

$$\sum_{j=1}^{m}\pi_j=1$$  

$$L(\pi_1,...,\pi_m,\lambda)=Q[\Psi|\Psi(k)]-\lambda(\sum_{j=1}^{m}\pi_j -1)=0$$   

Calculate the first derivative of $L(\pi_1,...,\pi_m,\lambda)$ respect to $\pi_j$ and $\lambda$, for $j=1,...,m$:  
$$L(\pi_1,...,\pi_m,\lambda)^{'}_{\pi_j}=0;  L(\pi_1,...,\pi_m,\lambda)^{'}_{\lambda}=0$$    

$$\sum_{i=1}^{n}p_{ij}^{(k+1)}(\cfrac{1}{\pi_j})-\lambda=0$$  

Slove for $\pi_j$
$$\pi_j=\cfrac{\sum_{i=1}^{n}p_{ij}^{(k+1)}}{\lambda}$$  

$$\sum_{j=1}^{m}\pi_j=\cfrac{\sum_{j=1}^{m}\sum_{i=1}^{n}p_{ij}^{(k+1)}}{\lambda}=\cfrac{\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}}{\lambda}=\cfrac{n}{\lambda}=1; \lambda=n$$  

$$\pi_j=\cfrac{\sum_{i=1}^{n}p_{ij}^{(k+1)}}{n}$$  


$$Q[\Psi|\Psi(k)]=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}[log(\pi_j)+log(\cfrac{1}{\sqrt{2\pi} \sigma}exp(-\cfrac{(y_i-x_i^T \beta_j)^2}{2\sigma^2}))]$$  
$$=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}[log(\pi_j)+log(\cfrac{1}{\sqrt{2\pi}\sigma})+(-\cfrac{(y_i-x_i^T \beta_j)^2}{2\sigma^2})]$$   
$$=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}log(\pi_j)-\cfrac{1}{2}\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)} log(2\pi\sigma^2)-\cfrac{1}{2}\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}\cfrac{(y_i-x_i^T \beta_j)^2}{\sigma^2}=I_1-\cfrac{1}{2}I_2-\cfrac{1}{2}I_3$$

Where $I_1=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}log(\pi_j)$;$I_2=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)} log(2\pi\sigma^2)$; and $I_3=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}\cfrac{(y_i-x_i^T \beta_j)^2}{\sigma^2}$

$$I_3=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}(\cfrac{(y_i-x_i^T \beta_j)^2}{2\sigma^2})$$  
Which is a qudratic form. It is the sum of m quadratic forms each involving a single $\beta_j$ for every certain j.  

$$\sum_{i=1}^{n} p_{ij}^{(k+1)}(\cfrac{(y_i-x_i^T \beta_j)^2}{2\sigma^2})=\sum_{i=1}^{n}p_{ij}^{(k+1)}[x_i^T(\cfrac{y_i}{x_i^T}-\beta_j)^2]$$  

From the property of sample mean, $\beta_j$ must be the mean of a weighted sample of $\cfrac{y_1}{x_1},...,\cfrac{y_n}{x_n^T}$. And for each element, it has the weight: $p_{ij}\cdot x_i\cdot x_i^T$. It can be represented in the following form:

$$\sum_{i=1}^{n} p_{ij}^{(k+1)}x_ix_i^T(\cfrac{y_i}{x_i^T}-\beta_j)^2$$  

Solve for $\beta_j$:  
$$\beta_j=\cfrac{\sum_{i=1}^{n} p_{ij}^{(k+1)}x_ix_i^T(\cfrac{y_i}{x_i^T})}{\sum_{i=1}^{n} p_{ij}^{(k+1)}x_ix_i^T}=\cfrac{\sum_{i=1}^{n} p_{ij}^{(k+1)}x_i{y_i}}{\sum_{i=1}^{n} p_{ij}^{(k+1)}x_ix_i^T}$$  

Only $I_2$ and $I_3$ contain $\sigma$  

$$I_2+I_3=\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}ln(\sigma)+\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}\cfrac{(y_i-x_i^T \beta_j)^2}{2\sigma^2}$$  

Minimizing $I_2+I_3$ to maximaze $Q[\Psi|\Psi(k)]$. Now we knew that $\beta_j$ is the weighted mean of $\cfrac{y_i}{x_i^T}$, therefore, $\sigma$ must be the sample variance of all data for every $j=1,...,m$.  

$$\sigma^{2(k+1)}=\cfrac{\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2}{\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}}$$  

$$=\cfrac{\sum_{i=1}^{n} \sum_{j=1}^{m} p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2}{n}$$    

##Question2
###(a)


$$g(x)\propto(2x^{\theta-1}+x^{\theta-\frac{1}{2}}e^{-x})$$

$$C\int_0^\infty(2x^{\theta-1}+x^{\theta-\frac{1}{2}})e^{-x}dx = 1$$

$$2C\gamma_{(\theta)}+C\gamma_{(\theta+\frac{1}{2})} = 1$$

$$C = \frac{1}{2\gamma_{(\theta)}+C\gamma_{(\theta+\frac{1}{2})}}$$

$$\therefore g(x) = \frac{2\gamma_{(\theta)}}{2\gamma_{(\theta)}+\gamma_{(\theta+\frac{1}{2})}}\frac{1}{\gamma_{(\theta)}}x^{\theta-1}e^{-x}+\frac{\gamma_{(\theta+\frac{1}{2})}}{2\gamma_{(\theta)}+\gamma_{(\theta+\frac{1}{2})}}*\frac{1}{\gamma_{(\theta+\frac{1}{2})}}x^{\theta-\frac{1}{2}}e^{-x}$$


  It is the mixture of $\gamma(\theta,1)$ and $\gamma(\theta+\frac{1}{2},1)$, where the weights are $\frac{2\gamma_{(\theta)}}{2\gamma_{(\theta)}+\gamma_{(\theta+\frac{1}{2})}}$ and $\frac{\gamma_{(\theta+\frac{1}{2})}}{2\gamma_{(\theta)}+\gamma_{(\theta+\frac{1}{2})}}$

###(b)

Repeat the following step 1 and step 2 for 10,000 times:  
**Step 1**: $u$ ~ $unif (0,1)$  
**Step 2**: find $x$, where $Cg(x)=u$  
            $sample_g \leftarrow c (sample_g, x)$  
            
```{r}
#you can change the value of theta, we set it equals to 2 in order to calculate conveniently
theta <- 2
C <- 1 / (2 * gamma(theta) + gamma(theta + 1/2))

#pdf of g(x)
func_g <- function(x){ 
  return(C * (2 * x ** (theta - 1) + x ** (theta - 1/2)) * exp(-x))
}

#calculate the inverse of g-unit(1)
inverse_function <- function(x){
  #u <- runif(1)
  cdf_g <- integrate(func_g, lower = 0, upper = x)$value
  inverse_value <- cdf_g - u
  return(inverse_value)
  #return()
}

#generate 10000 samples
sample_g <- c()
for(i in 1:10000){
  u <- runif(1)
  inverse_method <- uniroot(inverse_function,interval = c(0,20))$root
  sample_g <- c(sample_g,inverse_method)
}

sampleg_pdf <- density(sample_g)

#calculate the true value of g_pdf
x <- seq(from = 0, to = 10, by = 0.0001)
g_value <- sapply(x,func_g)

##--------------Q2-b----------------------
```   

```{r}
#plot the true density and the sample density in one figure
plot(g_value~x,type = "l", xlab = "x", ylab = "density")
lines(sampleg_pdf, col = "red")
#add legend the fist two parameters denote the location, lty =1 denotes the type of legend is line
legend(6,0.33,c("true density of g", "sample density"), col = c("black","red"),text.col = c("black","red"),lty = c(1,1))
```

###(c)  

$\sqrt{4+x}<2+\sqrt{x}$ for all $x>0$ because $(2+\sqrt{x})^2-(\sqrt{4+x})^2=4\cdot \sqrt{x}>0$.   
Therefore, $\cfrac{f(x)}{g(x)}\leq1$, we choose $\alpha=\frac{1}{C}$.  

**Step 1**: $u$ ~ $unif (0, 1)$  
**Step 2**: find $x^*$, where $Cg(x)=u$  
            Calculate $f(x^*)$ and $g(x^*)$   
            $u^*$ ~ $unif(0,1)$  
            if $u^*<\cfrac{f(x^*)}{g(x^*)}$  
            $sample \leftarrow c(sample, x^*)$  
Repeat Step 1 and Step 2 until the length of sample is 10,000  

```{r}
#you can change the value of theta, we set it equals to 2 in order to calculate conveniently
theta <- 2
C <- 1 / (2 * gamma(theta) + gamma(theta + 1/2))

#pdf of g(x)
func_g <- function(x){ 
  return(C * (2 * x ** (theta - 1) + x ** (theta - 1/2)) * exp(-x))
}

#calculate the inverse of g-unit(1)
inverse_function <- function(x){
  #u <- runif(1)
  cdf_g <- integrate(func_g, lower = 0, upper = x)$value
  inverse_value <- cdf_g - u
  return(inverse_value)
  #return()
}
##--------------Q2-c-----------------------
```  

```{r}
#pdf of f
func_f <- function(x){
  return(((4 + x) ** 0.5) * (x ** (theta - 1)) * exp(- x))
}

const <- 1 / integrate(func_f,lower = 0, upper = Inf)$value
func_f_true <- function(x){
  value <- func_f(x) * const
  return(value)
}
#calculate the true value of f_pdf
x <- seq(from = 0, to = 10, by = 0.0001)
f_value <- sapply(x,func_f_true)

f_sample_rejection <-c() 
while(length(f_sample_rejection)<10000){
  u <- runif(1)
  sample_value <- uniroot(inverse_function,interval = c(0,20))$root
  sample_value_g <- func_g(sample_value) / C
  sample_value_f <- func_f(sample_value)
  ratio <- sample_value_f / sample_value_g
  rejection_value <- runif(1)
  if(ratio >= rejection_value){
    f_sample_rejection <- c(f_sample_rejection, sample_value)
  }
}

sampleg_pdf <- density(f_sample_rejection)
##--------------Q2-c----------------------
```  

```{r}
#plot the true density and the sample density in one figure
plot(f_value~x,type = "l", xlab = "x", ylab = "density")
lines(sampleg_pdf, col = "red")
#add legend the fist two parameters denote the location, lty =1 denotes the type of legend is line
legend(6,0.33,c("true density of f", "sample density"), col = c("black","red"),text.col = c("black","red"),lty = c(1,1))
```  


##Question 3  
###(a)  

For$$0<x<1$$,

$$f(x)=\cfrac{x^{\theta-1}}{1+x^2}+\sqrt{2+x^2}\cdot (1-x)^{\beta-1} \leq x^{\theta -1}+\sqrt{3}\cdot (1-x)^{\beta-1}$$

  

$$g^*_1(x)=x^{\theta-1}\\ \int^1_0g^*_1(x)dx=\cfrac{x^{\theta}}{\theta}|^1_0=\cfrac{1}{\theta}=\cfrac{1}{C_1}\\ \Rightarrow g_1(x)=\theta\cdot x^{\theta-1}$$  

$$g^*_2(x)=\sqrt{3}\cdot (1-x)^{\beta-1}\\ \int^1_0g^*_2(x)dx=-\sqrt{3}\cdot \cfrac{(1-x)^\beta}{\beta}|^1_0=\cfrac{\sqrt3}{\beta}=\cfrac{1}{C_2}\\ \Rightarrow g_2(x)=\beta\cdot (1-x)^{\beta-1}$$   

$$g(x)=\cfrac{1}{2}g_1(x)+\cfrac{1}{2}g_2(x)$$  

$$f(x)\leq x^{\theta-1}+\sqrt{3}\cdot (1-x)^{\beta-1}$$

$$\alpha g(x)=\alpha\cdot \cfrac{1}{2}\cdot \theta\cdot x^{\theta-1}+\alpha\cdot \cfrac{1}{2}\cdot \beta\cdot (1-x)^{\beta-1}$$
$$1\leq \cfrac{1}{2}\cdot \alpha\cdot\theta$$

$$\cfrac{2}{\theta}\leq \alpha$$

$$1\leq \cfrac{1}{2}\cdot \alpha\cdot\beta$$
$$\cfrac{2}{\beta}\leq \alpha$$  

$$\therefore \alpha = max(\cfrac{2}{\theta}, \cfrac{2}{\beta})$$  

**Step 1**: $u$-weight ~ $unif(0,1)$  
**Step 2**: if $u$-weight $\leq w_1$  
            $u$ ~ $unif(0,1)$, find $x^*$, where $g_1(x)=u$  
            elase,   
            $u$ ~ $unif(0,1)$, find $x^*$, where $g_2(x)=u$   
**Step 3**: calculate $f(x^*)$ and $g(x^*)$  
            $u^*$ ~ $unif(0,1)$  
            if $u^* \leq \cfrac{f(x^*)}{g(x^*)}$  
            $sample \leftarrow c (sample, x^*)$  
Repeat Step 1,2 and 3 until the length of sample is 10,000 
```{r}
#you can change the value of theta and beta
theta_factor <- 9
beta_factor <- 6
# W1 <- (C1^2) / (C1 + C2)#weight for g1(x)
# W2 <- (C2^2) / (C1 + C2)#weight for g2(x)
W1 <- 1/2#weight for g1(x)
W2 <- 1/2#weight for g2(x)
alpha_factor <- max(2/theta_factor, 2/beta_factor)

#pdf of g(x)
func_g1 <- function(x){ 
  g1 <- theta_factor * (x ^ (theta_factor-1))
  return(g1)
}

func_g2 <- function(x){ 
  g2 <- (beta_factor) * ((1-x)^(beta_factor-1))
  return(g2)
}

func_g_mix <- function(x){ 
  g1 <- func_g1(x)
  g2 <- func_g2(x)
  return(W1*g1+W2*g2)
}
# x <- seq(from = 0, to = 1, by = 0.001)
# cdf_g <- c()
# for(i in 1:length(x)){
#   cdf_g <- c(cdf_g, integrate(func_g, lower = 0, upper = x[i])$value)
# }
# #g_value <- sapply(x,func_g)
# plot(cdf_g~x,type = "l", xlab = "x", ylab = "density")
#calculate the inverse of g-unit(1)
inverse_function <- function(x){
  #u <- runif(1)
  cdf_g <- integrate(func_g, lower = 0, upper = x)$value
  inverse_value <- cdf_g - u
  return(inverse_value)
}
##--------------Q3-a-----------------------
#pdf of f
func_f <- function(x){
  f1 <- (x ^ (theta_factor-1)) / (1+x^2)
  f2 <- ((2+x^2)^0.5) * ((1-x)^(beta_factor - 1))
  return(f1+f2)
}
func_f_true <- function(x){
  value <- func_f(x) * const
  return(value)
}
#solve for the constant for f_pdf
integrate(func_f, lower = 0, upper = 1)$value
const <- 1/integrate(func_f, lower = 0, upper = 1)$value
#true value of f_pdf
x <- seq(from = 0, to = 1, by = 0.001)
f_value <- sapply(x,func_f_true)

f_sample_rejection <-c() 
while(length(f_sample_rejection)<10000){
  u_weight <- runif(1)
  if(u_weight < 0.5){
    u <- runif(1)
    func_g <- func_g1
    sample_value <- uniroot(inverse_function,interval = c(0,1))$root
  }
  else{
    u <- runif(1)
    func_g <- func_g2
    sample_value <- uniroot(inverse_function,interval = c(0,1))$root
  }
  sample_value_g <- func_g_mix(sample_value) 
  sample_value_f <- func_f(sample_value)
  ratio <- sample_value_f / (alpha_factor * sample_value_g)
  rejection_value <- runif(1)
  if(ratio >= rejection_value){
    f_sample_rejection <- c(f_sample_rejection, sample_value)
  }
}

sample_pdf <- density(f_sample_rejection)

#plot the true density and the sample density in one figure
plot(f_value~x,type = "l",col = "red")
lines(sample_pdf,type = "l", xlab = "x", ylab = "density", main = "using rejection method to get sample from f")
legend(0.6,4.5,c("sample density", "true density of f"), col = c("black","red"),text.col = c("black","red"),lty = c(1,1))
# str(f_sample_rejection)
# max(f_sample_rejection)
```


###(b)   


$$f(x_1)=\cfrac{x^{\theta-1}}{1+x^2} \leq x^{\theta-1}=\alpha_1\cdot g_1(x)$$  

$$f(x_2)=\sqrt{2+x^2}\cdot (1-x)^{\beta-1} \leq \sqrt{3}\cdot (1-x)^{\beta-1}=\alpha_2\cdot g_2(x)$$  

Then, $\alpha_1=\cfrac{1}{\theta}$, and $\alpha_2=\cfrac{\sqrt{3}}{\beta}$  

Thus, when $runif>\cfrac{\cfrac{1}{\theta}}{\cfrac{1}{\theta}+\cfrac{\sqrt{3}}{\beta}}$, sample from $g_2(x)$  

**Step 1**: $u$-weight ~ $unif(0,1)$

**Step 2**: if $u$-weight > weighted-boundary  
            $u$ ~ $unif(0,1)$, find $x^*$, where $g_2(x^*)=u$  
            calculate ratio $=\cfrac{f_2(x^*)}{\alpha g_2(x^*)}$  
            else,  
            $u$ ~ $unif(0,1)$, find $x^*$, where $g_1(x^*)=u$  
            calculate ratio $=\cfrac{f_1(x^*)}{\alpha g_1(x^*)}$
            
**Step 3**: $u^*$ ~ $unif(0,1)$   
            if $u^*<ratio$  
            $sample \leftarrow c (sample, x^*)$

Repeat Step 1, 2 and 3 until the length of sample is 10,000 
```{r}
#you can change the value of theta and beta
theta_factor <- 6
beta_factor <- 3
alpha_1 <- 1 / theta_factor
alpha_2 <- (3^0.5) / beta_factor

weight_boundary <-alpha_1 / (alpha_1 + alpha_2) 
#pdf of g(x)
func_g1 <- function(x){ 
  g1 <- theta_factor * (x ^ (theta_factor-1))
  return(g1)
}
func_g2 <- function(x){ 
  g2 <- (beta_factor) * ((1-x)^(beta_factor-1))
  return(g2)
}

# x <- seq(from = 0, to = 1, by = 0.001)
# cdf_g <- c()
# for(i in 1:length(x)){
#   cdf_g <- c(cdf_g, integrate(func_g, lower = 0, upper = x[i])$value)
# }
# #g_value <- sapply(x,func_g)
# plot(cdf_g~x,type = "l", xlab = "x", ylab = "density")
#calculate the inverse of g-unit(1)
inverse_function <- function(x){
  cdf_g <- integrate(func_g, lower = 0, upper = x)$value
  inverse_value <- cdf_g - u
  return(inverse_value)
}
##--------------Q3-b-----------------------
#pdf of f
func_f1 <- function(x){
  f1 <- (x ^ (theta_factor-1)) / (1+x^2)
  return(f1)
}

func_f2 <- function(x){
  f2 <- ((2+x^2)^0.5) * ((1-x)^(beta_factor - 1))
  return(f2)
}
func_f <- function(x){
  f1 <- func_f1(x)
  f2 <- func_f2(x)
  return(f1+f2)
}
integrate(func_f, lower = 0, upper = 1)$value
const <- 1/integrate(func_f, lower = 0, upper = 1)$value

func_f_true <- function(x){
  value <- func_f(x) * const
  return(value)
}
#solve for the constant for f_pdf

#true value of f_pdf
x <- seq(from = 0, to = 1, by = 0.001)
f_value <- sapply(x,func_f_true)
f_sample_rejection <-c() 
while(length(f_sample_rejection)<10000){
  random_weight <- runif(1)
  if(random_weight > weight_boundary){
    u <- runif(1)
    func_g <- func_g2
    sample_value <- uniroot(inverse_function,interval = c(0,1))$root
    sample_value_g <- func_g2(sample_value) 
    sample_value_f <- func_f2(sample_value)
    ratio <- sample_value_f / (alpha_2 * sample_value_g)
  }
  else{
    u <- runif(1)
    func_g <- func_g1
    sample_value <- uniroot(inverse_function,interval = c(0,1))$root
    sample_value_g <- func_g1(sample_value) 
    sample_value_f <- func_f1(sample_value)
    ratio <- sample_value_f / (alpha_1 * sample_value_g)
  }
  rejection_value <- runif(1)
  if(ratio >= rejection_value){
    f_sample_rejection <- c(f_sample_rejection, sample_value)
  }
}

sample_pdf <- density(f_sample_rejection)

#plot the true density and the sample density in one figure
plot(f_value~x,type = "l",col = "red")
lines(sample_pdf,type = "l", xlab = "x", ylab = "density", main = "using rejection method to get sample from f")
legend(0.6,2.3,c("sample density", "true density of f"), col = c("black","red"),text.col = c("black","red"),lty = c(1,1))
# str(f_sample_rejection)
# max(f_sample_rejection)
```
